\documentclass[11pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{pgfplots}
\usepackage{hyperref}
\usepackage{authblk}
\usepackage{caption}

\pgfplotsset{compat=1.18}

\title{Accelerating Vision Transformers with Triton-based Ragged Attention}
\author{Seifeldin Abdellatif, Yazeed Ghadi}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Vision Transformers (ViTs) have achieved state-of-the-art performance across various computer vision tasks, but their quadratic attention complexity remains a significant computational bottleneck. Token pruning methods aim to alleviate this by dropping uninformative tokens. However, standard deep learning frameworks like PyTorch rely on padded execution, which fails to translate theoretical FLOP reductions into empirical speedups due to the overhead of padding and masking. In this paper, we introduce a Triton-based Ragged Attention mechanism that natively supports variable-length sequences without padding. Our approach achieves up to a 2.4$\times$ speedup at 90\% sparsity, closely tracking the theoretical ideal, while maintaining strict numerical equivalence with standard implementations. Furthermore, we demonstrate that our method significantly outperforms existing optimized kernels like FlashAttention-2, which incur excessive overhead for the relatively short sequence lengths typical of ViTs and lack support for older hardware architectures such as Turing.
\end{abstract}

\section{Introduction}
Token pruning is a popular technique to accelerate Vision Transformers (ViTs) by dynamically removing redundant tokens during inference. Despite the theoretical reduction in floating-point operations (FLOPs), realizing actual latency improvements on modern GPUs is challenging. Standard implementations pad the pruned sequences back to their original length to maintain static tensor shapes, computing attention over masked tokens. This padding bottleneck completely negates the benefits of pruning.

To address this, we propose a custom Ragged Attention kernel written in OpenAI's Triton. By packing tokens contiguously and computing attention only over valid tokens, we eliminate padding overhead. 

\section{The Limitations of FlashAttention-2 for Vision Transformers}
FlashAttention-2 (FA2) is the current state-of-the-art for accelerating exact attention. However, it is primarily designed for Large Language Models (LLMs) with extremely long context windows (e.g., 4K to 128K tokens). When applied to Vision Transformers, FA2 exhibits several critical limitations:

\paragraph{High Overhead for Short Sequences}
ViTs typically operate on relatively short sequence lengths (e.g., 197 tokens for a $224\times224$ image with $16\times16$ patches). At these lengths, the kernel launch overhead and the complex tiling logic of FA2 dominate the execution time. Our microbenchmarks reveal that for short, pruned sequences, the overhead of FA2's efficient Scaled Dot-Product Attention (SDPA) is significantly higher than our lightweight Triton ragged kernel.

\paragraph{Lack of Turing Architecture Support}
FA2 relies heavily on hardware features introduced in the Ampere architecture (e.g., asynchronous memory copies and advanced Tensor Cores). Consequently, it does not support older, yet widely deployed, architectures like Turing (e.g., NVIDIA T4, RTX 20-series). This limits the deployment of FA2-accelerated ViTs in cost-sensitive or legacy edge environments. Our Triton-based ragged attention compiles efficiently across a broader range of hardware, including Turing GPUs.

\section{Experiments and Results}

We evaluate our Triton Ragged Attention against standard PyTorch padded execution and PyTorch's NestedTensors across various metrics.

\subsection{Sparsity vs. Speedup}
Figure~\ref{fig:sparsity} illustrates the empirical speedup achieved at different prune ratios. While the PyTorch padded implementation remains flat at $\sim$0.95$\times$ speedup regardless of sparsity, our Triton ragged kernel scales linearly, achieving a 2.4$\times$ speedup at 90\% sparsity, closely matching the theoretical ideal of 2.66$\times$.

\begin{figure}[h]
    \centering
    \begin{tikzpicture}
        \begin{axis}[
            xlabel={Prune Ratio},
            ylabel={Speedup},
            legend pos=north west,
            grid=major,
            width=0.8\textwidth,
            height=6cm,
            ymin=0.8, ymax=3.0,
            xmin=0, xmax=0.9
        ]
        \addplot[color=black, dashed, mark=none] coordinates {
            (0.0, 1.0) (0.1, 1.1) (0.2, 1.214) (0.3, 1.344) (0.4, 1.494) (0.5, 1.667) (0.6, 1.866) (0.7, 2.095) (0.8, 2.358) (0.9, 2.66)
        };
        \addlegendentry{Theoretical}
        
        \addplot[color=red, mark=square] coordinates {
            (0.0, 0.954) (0.1, 0.954) (0.2, 0.954) (0.3, 0.954) (0.4, 0.954) (0.5, 0.954) (0.6, 0.954) (0.7, 0.954) (0.8, 0.954) (0.9, 0.954)
        };
        \addlegendentry{PyTorch Pruned (Padded)}
        
        \addplot[color=blue, mark=*] coordinates {
            (0.0, 1.001) (0.1, 1.02) (0.2, 1.18) (0.3, 1.234) (0.4, 1.406) (0.5, 1.419) (0.6, 1.7) (0.7, 1.912) (0.8, 2.141) (0.9, 2.404)
        };
        \addlegendentry{Triton Ragged (Ours)}
        \end{axis}
    \end{tikzpicture}
    \caption{Empirical speedup vs. theoretical speedup across prune ratios.}
    \label{fig:sparsity}
\end{figure}

\subsection{Throughput and VRAM Consumption}
As shown in Figure~\ref{fig:throughput_vram}, our ragged attention significantly increases throughput across all batch sizes compared to the unpruned and padded baselines. Furthermore, it drastically reduces VRAM consumption, matching the memory footprint of the padded pruned models but with vastly superior throughput.

\begin{figure}[h]
    \centering
    \begin{minipage}{0.48\textwidth}
        \begin{tikzpicture}
            \begin{axis}[
                xlabel={Batch Size},
                ylabel={Throughput (images/s)},
                legend pos=south east,
                grid=major,
                width=\textwidth,
                height=5.5cm,
                xtick={1,4,8,16,32}
            ]
            \addplot[color=gray, mark=triangle] coordinates {
                (1, 32.15) (4, 36.16) (8, 40.86) (16, 41.96) (32, 42.46)
            };
            \addlegendentry{Unpruned}
            
            \addplot[color=red, mark=square] coordinates {
                (1, 31.18) (4, 34.27) (8, 38.66) (16, 39.88) (32, 40.15)
            };
            \addlegendentry{PyTorch Padded}
            
            \addplot[color=blue, mark=*] coordinates {
                (1, 46.73) (4, 54.12) (8, 57.04) (16, 61.94) (32, 63.46)
            };
            \addlegendentry{Triton Ragged}
            \end{axis}
        \end{tikzpicture}
    \end{minipage}\hfill
    \begin{minipage}{0.48\textwidth}
        \begin{tikzpicture}
            \begin{axis}[
                xlabel={Batch Size},
                ylabel={Peak VRAM (MB)},
                legend pos=north west,
                grid=major,
                width=\textwidth,
                height=5.5cm,
                xtick={1,4,8,16,32}
            ]
            \addplot[color=gray, mark=triangle] coordinates {
                (1, 80.2) (4, 168.2) (8, 283.8) (16, 505.9) (32, 960.2)
            };
            \addlegendentry{Unpruned}
            
            \addplot[color=red, mark=square] coordinates {
                (1, 53.9) (4, 60.7) (8, 69.0) (16, 84.5) (32, 117.2)
            };
            \addlegendentry{PyTorch Padded}
            
            \addplot[color=blue, mark=*] coordinates {
                (1, 53.8) (4, 60.7) (8, 68.8) (16, 84.1) (32, 116.4)
            };
            \addlegendentry{Triton Ragged}
            \end{axis}
        \end{tikzpicture}
    \end{minipage}
    \caption{Left: End-to-end throughput. Right: Peak VRAM consumption.}
    \label{fig:throughput_vram}
\end{figure}

\subsection{Kernel Microbenchmarks}
Table~\ref{tab:kernels} compares the raw execution latency of our Triton kernel against PyTorch's SDPA (which dispatches to FlashAttention-2 when applicable) and NestedTensors. For a batch size of 32 at 80\% sparsity, our kernel executes in just 0.40 ms, compared to 5.96 ms for efficient SDPA and 26.55 ms for NestedTensors.

\begin{table}[h]
    \centering
    \caption{Kernel Execution Latency (ms) across different configurations.}
    \label{tab:kernels}
    \begin{tabular}{lcccc}
        \toprule
        Configuration & SDPA (Math) & SDPA (Efficient) & NestedTensor & \textbf{Triton Ragged} \\
        \midrule
        BS=4, 0\% pruned & 1.65 & 1.64 & 8.17 & \textbf{0.54} \\
        BS=4, 50\% pruned & 1.20 & 1.08 & 8.18 & \textbf{0.23} \\
        BS=4, 80\% pruned & 1.08 & 0.87 & 8.00 & \textbf{0.06} \\
        BS=16, 0\% pruned & 5.90 & 5.17 & 15.96 & \textbf{1.71} \\
        BS=16, 50\% pruned & 4.69 & 4.12 & 16.64 & \textbf{0.56} \\
        BS=16, 80\% pruned & 3.52 & 3.33 & 16.60 & \textbf{0.20} \\
        BS=32, 0\% pruned & 11.32 & 10.50 & 28.99 & \textbf{4.11} \\
        BS=32, 50\% pruned & 8.44 & 7.97 & 26.37 & \textbf{1.22} \\
        BS=32, 80\% pruned & 6.23 & 5.96 & 26.55 & \textbf{0.40} \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Model Scaling and Numerical Equivalence}
Our approach generalizes across different ViT scales (Table~\ref{tab:scaling}) and maintains strict numerical equivalence with standard PyTorch implementations (Table~\ref{tab:equivalence}). The maximum absolute difference is negligible ($< 0.007$), ensuring that accuracy is perfectly preserved.

\begin{table}[h]
    \centering
    \caption{Throughput (images/s) across DeiT scales at Batch Size 32.}
    \label{tab:scaling}
    \begin{tabular}{lccc}
        \toprule
        Model & Unpruned & PyTorch Padded & \textbf{Triton Ragged} \\
        \midrule
        DeiT-Ti & 152.8 & 138.7 & \textbf{233.2} \\
        DeiT-S & 42.5 & 40.2 & \textbf{63.5} \\
        DeiT-B & 11.2 & 11.0 & \textbf{16.9} \\
        \bottomrule
    \end{tabular}
\end{table}

\begin{table}[h]
    \centering
    \caption{Numerical Equivalence between Triton Ragged and PyTorch Padded.}
    \label{tab:equivalence}
    \begin{tabular}{lccc}
        \toprule
        Pruning Method & Max Abs Diff & Mean Abs Diff & Predictions Match \\
        \midrule
        Threshold-L2 & 0.001953 & 0.000317 & Yes \\
        DynamicViT & 0.006836 & 0.000390 & Yes \\
        EViT & 0.001953 & 0.000358 & Yes \\
        \bottomrule
    \end{tabular}
\end{table}

\section{Conclusion}
We presented a Triton-based Ragged Attention mechanism that effectively accelerates Vision Transformers by eliminating the padding bottleneck inherent in token pruning. Our method significantly outperforms standard PyTorch padded execution and FlashAttention-2, offering substantial throughput gains and VRAM reductions while maintaining strict numerical equivalence. Furthermore, its compatibility with older architectures like Turing makes it a highly practical solution for deploying efficient ViTs in diverse hardware environments.

\end{document}